---
title: "Quarto Document Day 2"
format: html
editor: visual
---

## Original Tutorial Link

<https://satijalab.org/seurat/articles/pbmc3k_tutorial.html>

# Seurat - Guided Clustering Tutorial

When you click the **Render** button a document will be generated that includes both content and the output of embedded code. You can embed code like this:

```{r}
library(Seurat)
library(tidyverse)
library(patchwork)
```

You can add options to executable code like this

```{r}
pbmc <- readRDS('~/R projects/day1-20250727T114252Z-1-001/day1/02_project_start/gex_only_ubds3.RDS')
pbmc
```

```{r}
# View gene names stored in the RNA assay
rownames(pbmc)
```

```{r}
# Lets examine a few genes in the first thirty cells
pbmc[c("Il1rl1", "Creb1", "Bcl2"), 1:30]
```

```{r}
# The [[ operator can add columns to object metadata. This is a great place to stash QC stats
#pbmc[["percent.mt"]] <- PercentageFeatureSet(pbmc, pattern = "^MT-")
#pbmc
```

```{r}
# Show QC metrics for the first 5 cells
head(pbmc@meta.data, 5)
```

```{r}
VlnPlot(pbmc, features = c("nFeature_RNA"), ncol = 1, pt.size = 0.1)
```

```{r}
VlnPlot(pbmc, features = c("nFeature_RNA"), ncol = 1, pt.size = 0)
```

```{r}
VlnPlot(pbmc, features = c("nCount_RNA"), ncol = 1, pt.size = 0.1)
```

```{r}
VlnPlot(pbmc, features = c("nCount_RNA"), ncol = 1, pt.size = 0)
```

```{r}
VlnPlot(pbmc, features = c("percent.mt"), ncol = 1, pt.size = 0.1)
```

```{r}
VlnPlot(pbmc, features = c("percent.mt"), ncol = 1, pt.size = 0)
```

```{r}
# FeatureScatter is typically used to visualize feature-feature relationships, but can be used
# for anything calculated by the object, i.e. columns in object metadata, PC scores etc.

#plot1 <- FeatureScatter(pbmc, feature1 = "nCount_RNA", feature2 = "percent.mt")

plot2 <- FeatureScatter(pbmc, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
plot2

#plot1 + plot2
```

Filtering — were issues with too harsh filtering

```{r}
#too harsh filtering
#qced_pbmc <- subset(pbmc, subset = nFeature_RNA > 100 & nFeature_RNA < 1000 & percent.mt < 5)
#qced_pbmc
#qced_pbmc <- subset(pbmc, subset = nFeature_RNA > 100 & nFeature_RNA < 1000 & percent.mt < 5)
#qced_pbmc
```

```{r}
qced_pbmc <- subset(pbmc, subset = nFeature_RNA > 42 & nFeature_RNA < 120 & nCount_RNA > 10 & nCount_RNA < 2000)

```

```{r}
qced_pbmc
```

```{r}
VlnPlot(qced_pbmc, features = c("nFeature_RNA"), ncol = 1, pt.size = 0.1)
```

```{r}
VlnPlot(qced_pbmc, features = c("nCount_RNA"), ncol = 1, pt.size = 0.1)
```

Number of cells per experiment after QC

```{r}
head(qced_pbmc@meta.data)
```

```{r}
table(qced_pbmc$orig.ident)
```

```{r}
plot3 <- FeatureScatter(qced_pbmc, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
plot3
```

Normalization

```{r}
norm_qced_pbmc <- NormalizeData(qced_pbmc, normalization.method = "LogNormalize", scale.factor = 10000)
```

## Identification of highly variable features (feature selection)

```{r}
high_variable_pbmc <- FindVariableFeatures(norm_qced_pbmc, selection.method = "vst", nfeatures = 2000)

# Identify the 10 most highly variable genes
top10 <- head(VariableFeatures(high_variable_pbmc), 10)

# plot variable features with and without labels
plotA <- VariableFeaturePlot(high_variable_pbmc)
plotB <- LabelPoints(plot = plotA, points = top10, repel = TRUE)
plotA + plotB
```

## Scaling the data

```{r}
all.genes <- rownames(pbmc)
scale_pbmc <- ScaleData(high_variable_pbmc, features = all.genes)
```

## Perform linear dimensional reduction

Next we perform PCA on the scaled data. By default, only the previously determined variable features are used as input, but can be defined using `features` argument if you wish to choose a different subset (if you do want to use a custom subset of features, make sure you pass these to `ScaleData` first).

```{r}
lin_dim_red_pbmc <- RunPCA(scale_pbmc, features = VariableFeatures(object = scale_pbmc))
lin_dim_red_pbmc
```

Seurat provides several useful ways of visualizing both cells and features that define the PCA, including `VizDimReduction()`, [`DimPlot()`](https://satijalab.org/seurat/reference/dimplot), and [`DimHeatmap()`](https://satijalab.org/seurat/reference/dimheatmap)

```{r}
# Examine and visualize PCA results a few different ways
print(lin_dim_red_pbmc[["pca"]], dims = 1:5, nfeatures = 5)
```

```{r}
VizDimLoadings(lin_dim_red_pbmc, dims = 1:2, reduction = "pca")
```

```{r}
DimPlot(lin_dim_red_pbmc, reduction = "pca") + NoLegend()
```

An alternative heuristic method generates an ‘Elbow plot’: a ranking of principle components based on the percentage of variance explained by each one ([`ElbowPlot()`](https://satijalab.org/seurat/reference/elbowplot) function). In this example, we can observe an ‘elbow’ around PC9-10, suggesting that the majority of true signal is captured in the first 10 PCs.

```{r}
ElbowPlot(lin_dim_red_pbmc)
```

In particular [`DimHeatmap()`](https://satijalab.org/seurat/reference/dimheatmap) allows for easy exploration of the primary sources of heterogeneity in a dataset, and can be useful when trying to decide which PCs to include for further downstream analyses. Both cells and features are ordered according to their PCA scores. Setting `cells` to a number plots the ‘extreme’ cells on both ends of the spectrum, which dramatically speeds plotting for large datasets. Though clearly a supervised analysis, we find this to be a valuable tool for exploring correlated feature sets.

```{r}
DimHeatmap(lin_dim_red_pbmc, dims = 1, cells = 500, balanced = TRUE)
```

```{r}
DimHeatmap(lin_dim_red_pbmc, dims = 1:15, cells = 500, balanced = TRUE)
```

## Cluster the cells

|     |                   |     |                                              |     |
|-----|-------------------|-----|----------------------------------------------|-----|
|     |                   |     |                                              |     |
|     | Step              |     | Purpose                                      |     |
|     |                   |     |                                              |     |
|     | `FindNeighbors()` |     | Create similarity graph between cells        |     |
|     |                   |     |                                              |     |
|     | `FindClusters()`  |     | Detect clusters (groups) from graph          |     |
|     |                   |     |                                              |     |
|     | `Idents`          |     | Stores current identity/cluster of each cell |     |
|     |                   |     |                                              |     |

```{r}
obj <- FindNeighbors(obj, dims = 1:10)
obj <- FindClusters(obj, resolution = 0.5)
```

-   **Modularity** is a measure of how well a network (graph) is divided into **clusters or communities**. It compares the actual density of edges within clusters to the expected density if the edges were random. A **higher modularity** means better-defined, more distinct clusters

```{r}
# Look at cluster IDs of the first 5 cells
head(Idents(obj), 5)
```

## Run non-linear dimensional reduction (UMAP/tSNE)

```{r}
obj <- RunUMAP(obj, dims = 1:12)
```

```{r}
# note that you can set `label = TRUE` or use the LabelClusters function to help label
# individual clusters
DimPlot(obj, reduction = "umap")
```

Plot the clusters that are computed on the same plot, paint by clusters

```{r}
saveRDS(obj, file = "~/DataProjectWeek2/hemato_preprocessed_nonlinred.rds")


```

```{r}
obj <- readRDS("~/DataProjectWeek2/hemato_preprocessed_nonlinred.rds")
```

```{r}
DimPlot(obj, reduction = "umap", group.by = "seurat_clusters")
```

## Finding differentially expressed features (cluster biomarkers)

```{r}
# find all markers of cluster 2
cluster2.markers <- FindMarkers(obj, ident.1 = 2)
head(cluster2.markers, n = 5)
```

```{r}
#how to go to different layers
obj@assays$RNA@layers$scale.data
```

```{r}
count_umap <- FeaturePlot(obj, reduction = "umap", features = "nCount_RNA")
feature_umap <- FeaturePlot(obj, reduction = "umap", features = "nFeature_RNA")
```

```{r}
FeaturePlot(obj, reduction = "umap", features = "nFeature_RNA")
```

```{r}
count_umap + feature_umap
```

## Finding differentially expressed features (cluster biomarkers)

```{r}
# find all markers of cluster 2
cluster2.markers <- FindMarkers(obj, ident.1 = 2)
head(cluster2.markers, n = 5)
```

```{r}
# find all markers distinguishing cluster 5 from clusters 0 and 3
cluster5.markers <- FindMarkers(obj, ident.1 = 5, ident.2 = c(0, 3))
head(cluster5.markers, n = 5)
```

1.  How does FindMarkers() function compare gene expression in different clusters
2.  why p_val and p_val_adj are 0

```{r}
# find markers for every cluster compared to all remaining cells, report only the positive
# ones
pbmc.markers <- FindAllMarkers(obj, only.pos = TRUE)
pbmc.markers %>%
    group_by(cluster) %>%
    dplyr::filter(avg_log2FC > 1)
```
