---
title: "30.07 Project Continue"
format: html
editor: visual
---

## Finding differentially expressed features (cluster biomarkers)

```{r}
obj <- readRDS("~/DataProjectWeek2/hemato_preprocessed_nonlinred.rds")
```

```{r}
test <- subset(obj, !(seurat_clusters %in% c(7, 11)))
```

```{r}
test
```

```{r}
#saving cleared dataset with removed clusters
saveRDS(test, file = "~/DataProjectWeek2/hemato_preprocessed_nonlinred_test.rds")
```

```{r}
#working with clean dataset
test_2 <- readRDS("~/DataProjectWeek2/hemato_preprocessed_nonlinred_test.rds")
```

```{r}
# find all markers of cluster 2
cluster2.markers <- FindMarkers(test_2, ident.1 = 2)
head(cluster2.markers, n = 5)
```

```{r}
# find all markers distinguishing cluster 5 from clusters 0 and 3
cluster5.markers <- FindMarkers(test_2, ident.1 = 5, ident.2 = c(0, 3))
head(cluster5.markers, n = 5)
```

1.  How does FindMarkers() function compares gene expression in different clusters
2.  why p_val and p_val_adj are 0

```{r}
# find markers for every cluster compared to all remaining cells, report only the positive ones
pbmc.markers <- FindAllMarkers(test_2, only.pos = TRUE)
pbmc.markers %>%
    group_by(cluster) %>%
    dplyr::filter(avg_log2FC > 1)
```

Test for differentiental expression

```{r}
cluster0.markers <- FindMarkers(test_2, ident.1 = 0, logfc.threshold = 0.25, test.use = "roc", only.pos = TRUE)
```

```{r}
VlnPlot(test_2, features = c("Mboat1", "Cyp11a1"))
```

```{r}
FeaturePlot(test_2, features = c("Mboat1", "Cyp11a1"))
```

```{r}
RidgePlot(test_2, features = c("Mboat1", "Cyp11a1"))
```

```{r}
DotPlot(test_2, features = c("Mboat1", "Cyp11a1"))
```

```{r}
# you can plot raw counts as well
VlnPlot(test_2, features = c("Mboat1", "Cyp11a1"), slot = "counts", log = TRUE)
```

```{r}
FeaturePlot(test_2, features = c("Rpa2", "Cks2", "Igfbp4", "Cyp11a1", "Prss34", "Impdh2", "Gmnn", "Slbp", "Kif23"))
```

```{r}
pbmc.markers %>%
    group_by(cluster) %>%
    dplyr::filter(avg_log2FC > 1) %>%
    slice_head(n = 10) %>%
    ungroup() -> top10
DoHeatmap(test_2, features = top10$gene) + NoLegend()
```

```{r}
# Get top 5 genes with lowest p-values per cluster
top_genes_per_cluster <- pbmc.markers %>%
  group_by(cluster) %>%
  arrange(p_val, .by_group = TRUE) %>%
  slice_head(n = 5)

top_genes_list <- split(top_genes_per_cluster$gene, top_genes_per_cluster$cluster)
top_genes_list

```

```{r}
top5 <- pbmc.markers %>%
  group_by(cluster) %>%
  dplyr::filter(avg_log2FC > 1) %>%
  arrange(p_val, .by_group = TRUE) %>%
  slice_head(n = 5) %>%
  ungroup()

DoHeatmap(test_2, features = top5$gene) + NoLegend()
```

```{r}
library(SingleR)
library(celldex)

ref <- celldex::HumanPrimaryCellAtlasData()
expr <- GetAssayData(test_2, slot = "data")
SingleR(test = expr, ref = ref, labels = ref$label.main)

# Assign predicted labels to Seurat object
obj$SingleR_labels <- pred$labels
Idents(test_2) <- obj$SingleR_labels

# Plot with annotated cell types
DimPlot(test_2, label = TRUE)
```

```{r}
library(Seurat)
library(patchwork)

resol_clust <- FindClusters(test_2, resolution = c(0.2, 0.6, 1.0))

resol_clust <- RunUMAP(resol_clust, dims = 1:20)

#resolution=0.2, has 8 clusters 

p1 <- DimPlot(resol_clust, group.by = "RNA_snn_res.0.2", reduction = "umap") + ggtitle("Resolution 0.2") + NoLegend()
p2 <- DimPlot(resol_clust, group.by = "RNA_snn_res.0.6", reduction = "umap") + ggtitle("Resolution 0.6") + NoLegend()
p3 <- DimPlot(resol_clust, group.by = "RNA_snn_res.1",   reduction = "umap") + ggtitle("Resolution 1.0") + NoLegend()

p1 + p2 + p3

```

```{r}
# Get top 5 genes with lowest p-values per cluster
top_genes_per_cluster_0.2 <- pbmc.markers %>%
    filter(cluster >= 0 & cluster <= 7)
 # only clusters 0 to 7 for resolution 0.2
  group_by(cluster) %>%
  arrange(p_val, .by_group = TRUE) %>%
  slice_head(n = 5)

top_genes_list_0.2 <- split(top_genes_per_cluster_0.2$gene, top_genes_per_cluster_0.2$cluster)
top_genes_list_0.2

```

```{r}
BiocManager::install("celldex")
BiocManager::install("SingleR")

library(SingleR)
library(celldex)

ref <- celldex::HumanPrimaryCellAtlasData()

expr <- GetAssayData(resol_clust, slot = "data")

singler.results <- SingleR(test = expr, ref = ref, labels = ref$label.main)

# Add SingleR labels to metadata
obj$SingleR.labels <- singler.results$labels

# Visualize
DimPlot(resol_clust, group.by = "SingleR.labels", label = TRUE) + NoLegend()

```

```{r}
colnames(colData(ref))
```

```{r}

install.packages("remotes")

# Install if not yet installed
if (!require("remotes")) install.packages("remotes")
remotes::install_github("IanevskiAleksandr/scType")

# Load libraries
library(scType)
library(Seurat)
library(dplyr)

# Load marker file for scType (human or mouse)
marker_file_url <- "https://raw.githubusercontent.com/IanevskiAleksandr/sc-type/master/ScType/data/ScTypeDB_full.csv"
markers_df <- read.csv(marker_file_url)

# Select only human markers
markers_human <- markers_df %>% filter(Species == "Human")

# Extract expression data from your Seurat object
expr_matrix <- GetAssayData(resol_clust, slot = "data")  # Use "counts" if needed

# Get cluster identities
cluster_ids <- Idents(obj)

# Run scType
sctype_results <- sctype(exprs = expr_matrix,
                         markers = markers_human,
                         clusters = cluster_ids,
                         tissue = "Immune system")  # or use "All" for broader match

# View top annotations per cluster
print(sctype_results$results$best)

# Rename cluster identities in the Seurat object
new_ids <- sctype_results$results$best$cell_type
names(new_ids) <- sctype_results$results$best$cluster
resol_clust <- RenameIdents(resol_clust, new_ids)

# Visualize UMAP with cell type labels
DimPlot(resol_clust, reduction = "umap", label = TRUE) + NoLegend()
```

1.  Google what genes correspond to what cell types
2.  Create a dictionary, table or seurat object to annotate genes, markers and cell types
3.  Recompute the differentially expressed genes and try to annotate the clusters with cell types

```{r}
{
  "Cluster 0": {
    "Genes": ["Slc6a4", "Hcp t8", "Ar14c", "Mboat1"],
    "Inferred Cell Type": "Neuronal Subtype or Specialized Cells"
  },
  "Cluster 1": {
    "Genes": ["Top2a", "Tacc3", "Cenpe", "Aurkb", "Tipin"],
    "Inferred Cell Type": "Proliferating Cells (High Expression of Cell Cycle Genes)"
  },
  "Cluster 2": {
    "Genes": ["Prss34", "Hcpt8", "Hgf", "Aqp9", "Cd200r3"],
    "Inferred Cell Type": "Neutrophils / Specialized Epithelial Cells"
  },
  "Cluster 3": {
    "Genes": ["Wfdc17", "Clec4n", "Mrc1", "Tgfbi", "Atp6v0d2"],
    "Inferred Cell Type": "Macrophages / Myeloid Cells"
  },
  "Cluster 4": {
    "Genes": ["Ctss", "Tgfbi", "Gda", "Hp", "Mpeg1"],
    "Inferred Cell Type": "Macrophages / Myeloid Cells"
  },
  "Cluster 5": {
    "Genes": ["Lrg1", "Hp", "C3", "Mgst2", "Gda"],
    "Inferred Cell Type": "Epithelial Cells or Liver Cells"
  },
  "Cluster 6": {
    "Genes": ["Rab38", "Slc18a2", "Car2", "Nkg7", "Impdh2"],
    "Inferred Cell Type": "Natural Killer (NK) Cells / Cytotoxic T Cells"
  },
  "Cluster 7": {
    "Genes": ["Cma1", "Gzmb", "Hbb-bt", "Prss34", "Kit"],
    "Inferred Cell Type": "Natural Killer (NK) Cells / Cytotoxic T Cells"
  },
  "Cluster 8": {
    "Genes": ["Cma1", "Gzmb", "Hbb-bt", "Prss34", "Kit"],
    "Inferred Cell Type": "Natural Killer (NK) Cells / Cytotoxic T Cells"
  },
  "Cluster 9": {
    "Genes": ["S100a9", "Lrg1", "Chi13", "Itgb2l", "Hp"],
    "Inferred Cell Type": "Neutrophils / Myeloid Cells"
  },
  "Cluster 10": {
    "Genes": ["Mpeg1", "Tgfbi", "Gpnmb", "Il6st", "Mrc1"],
    "Inferred Cell Type": "Macrophages"
  },
  "Cluster 11": {
    "Genes": ["Wfdc17", "Clec4n", "Gpr137b", "Tgfbi", "Mrc1"],
    "Inferred Cell Type": "Macrophages / Dendritic Cells"
  },
  "Cluster 12": {
    "Genes": ["Wfdc17", "Clec4n", "Gpr137b", "Tgfbi", "Mrc1"],
    "Inferred Cell Type": "Macrophages / Dendritic Cells"
  },
  "Cluster 13": {
    "Genes": ["Slc14a1", "Ehd3", "Slamf1", "Serpine2"],
    "Inferred Cell Type": "Erythroid Cells / Reticulocytes"
  }
}
```

```{r}
# Example data frame with Cluster IDs
df <- data.frame(
  Cell_ID = paste0("Cell", 1:20),
  Cluster = sample(0:13, 20, replace = TRUE)
)

# Your legend: cluster ID to inferred cell type mapping
cluster_annotation <- c(
  "0" = "Neuronal Subtype or Specialized Cells",
  "1" = "Proliferating Cells (High Expression of Cell Cycle Genes)",
  "2" = "Neutrophils / Specialized Epithelial Cells",
  "3" = "Macrophages / Myeloid Cells",
  "4" = "Macrophages / Myeloid Cells",
  "5" = "Epithelial Cells or Liver Cells",
  "6" = "Natural Killer (NK) Cells / Cytotoxic T Cells",
  "7" = "Natural Killer (NK) Cells / Cytotoxic T Cells",
  "8" = "Natural Killer (NK) Cells / Cytotoxic T Cells",
  "9" = "Neutrophils / Myeloid Cells",
  "10" = "Macrophages",
  "11" = "Macrophages / Dendritic Cells",
  "12" = "Macrophages / Dendritic Cells",
  "13" = "Erythroid Cells / Reticulocytes"
)

# Annotate the data frame using the cluster_annotation vector
df$Inferred_Cell_Type <- cluster_annotation[as.character(df$Cluster)]

# View the annotated data frame
print(df)
```

```{r}
library(Seurat)
library(patchwork)

# Your existing code:
resol_clust <- FindClusters(test_2, resolution = c(0.2))
#resol_clust <- RunUMAP(resol_clust, dims = 1:8)

p1 <- DimPlot(resol_clust, group.by = "RNA_snn_res.0.2", reduction = "umap") + ggtitle("Resolution 0.2") + NoLegend()

p1

# --------- Manual annotation for clusters at resolution 1.0 ----------

# Create a named vector for cluster-to-cell-type mapping (as characters)
cluster_annotation <- c(
  "0" = "Neuronal Subtype or Specialized Cells",
  "1" = "Proliferating Cells (High Expression of Cell Cycle Genes)",
  "2" = "Neutrophils / Specialized Epithelial Cells",
  "3" = "Macrophages / Myeloid Cells",
  "4" = "Macrophages / Myeloid Cells",
  "5" = "Epithelial Cells or Liver Cells",
  "6" = "Natural Killer (NK) Cells / Cytotoxic T Cells",
  "7" = "Natural Killer (NK) Cells / Cytotoxic T Cells",
  "8" = "Natural Killer (NK) Cells / Cytotoxic T Cells",
  "9" = "Neutrophils / Myeloid Cells",
  "10" = "Macrophages",
  "11" = "Macrophages / Dendritic Cells",
  "12" = "Macrophages / Dendritic Cells",
  "13" = "Erythroid Cells / Reticulocytes"
)

# Extract cluster IDs as character vector from metadata
clusters <- as.character(resol_clust@meta.data$RNA_snn_res.1)

# Add new metadata column for inferred cell types
resol_clust@meta.data$Inferred_Cell_Type <- cluster_annotation[clusters]

# Check annotation on UMAP plot with labels for cell types
DimPlot(resol_clust, group.by = "Inferred_Cell_Type", reduction = "umap") + ggtitle("Cell Type Annotation at Resolution 1.0")
```

```{r}
# Create a named vector with cluster IDs as names and cell types as values.
cluster_annotation <- c(
    "0" = "Neuronal Subtype",
    "1" = "Proliferating Cells",
    "2" = "Neutrophils",
    "3" = "Macrophages",
    "4" = "Macrophages",
    "5" = "Epithelial Cells",
    "6" = "NK/T Cells",
    "7" = "NK/T Cells",
    "8" = "NK/T Cells",
    "9" = "Neutrophils",
    "10" = "Macrophages",
    "11" = "Macrophages/Dendritic Cells",
    "12" = "Macrophages/Dendritic Cells",
    "13" = "Erythroid Cells"
)

# Now, add this new metadata column to your Seurat object.
# Assuming your main Seurat object is named 'obj'
obj$Inferred_Cell_Type <- cluster_annotation[obj$seurat_clusters]

# Now you can visualize the plot using the new labels.
DimPlot(obj,
        group.by = "Inferred_Cell_Type",
        reduction = "umap",
        label = TRUE,
        pt.size = 0.5)
```
